---
title: "Citation Snowballing for Literature Discovery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Citation Snowballing for Literature Discovery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
```

A keyword search only finds papers that use your keywords; it misses related work phrased in different terms. *Citation snowballing* follows citation links outward from a set of seed papers to surface those neighbors regardless of vocabulary -- a standard supplementary search step in evidence synthesis.

This vignette expands a seed corpus with `citation_snowball()`, inspects why each paper was admitted, and characterizes the *expansion space* (the citation-adjacent literature the query never returned) with MeSH keyness. Both `citation_snowball()` and `mesh_keyness()` are part of `puremoe`'s local analysis layer: transforms over tables `get_records()` already returned, with no further API calls. It proposes candidates to screen; it does not replace manual review. Note that iCite links cover PubMed-indexed articles only, so snowballing inherits PubMed's coverage.

```{r libs}
library(puremoe)
library(dplyr)
library(DT)
```

## Seed corpus

Search PubMed, then pull iCite records for the hits. Snowballing uses the `icites` endpoint specifically: each record carries a `citation_net` of that paper's references and citing papers.

```{r search}
pmids <- search_pubmed('"political ideology"[TiAb]')

length(pmids)

seed_icites <- get_records(pmids, endpoint = "icites", cores = 1L, sleep = 0.25)
```

## Expand by snowballing

`citation_snowball()` walks the links already in the iCite response (no extra API call). `direction = "both"` looks backward (papers the seeds cite) and forward (papers that cite the seeds); `min_links` admits a candidate only if it connects to at least that many seeds.

```{r snowball}
snowball <- seed_icites |>
  citation_snowball(direction = "both", min_links = 2)

snowball |>
  count(seed)   # seeds vs newly discovered candidates
```

## The audit trail

Every row carries its provenance: `seed`, `cited_links` (seeds that cite it), `citing_links` (seeds it cites), and `link_count` (the ranking total and `min_links` gate). Candidates are, by construction, papers the keyword query did not return.

### Ranked candidates

```{r audit}
candidates <- snowball |>
  filter(!seed) |>
  arrange(desc(link_count))

candidates |>
  head(25) |>
  DT::datatable(rownames = FALSE, options = list(scrollX = TRUE))
```

## The expansion space

Fetch metadata -- including MeSH -- once for the whole snowballed corpus (seeds plus candidates) and reuse it below. Joining the candidates back to their titles shows what the snowball surfaced.

### Snowballed candidates with metadata

```{r corpus-meta}
corpus_meta <- snowball$pmid |>
  get_records(endpoint = "pubmed_abstracts", cores = 1L, sleep = 0.25)

corpus_meta |>
  left_join(snowball, by = "pmid") |>
  filter(!seed) |>
  arrange(desc(link_count)) |>
  select(pmid, year, journal, articletitle,
         cited_links, citing_links, link_count) |>
  head(25) |>
  DT::datatable(rownames = FALSE, options = list(scrollX = TRUE))
```

### Keyness: how the corpus profile shifts

Raw MeSH counts are dominated by terms common everywhere (`Humans`). `mesh_keyness()` compares each descriptor's document rate in a corpus to its PubMed-wide rate from `data_mesh_frequencies`, returning log-odds or signed G2 scores. Here we use the default log-odds score and inspect the over-represented descriptors.

The seed corpus alone should surface the obvious topic terms, a sanity check that keyness behaves.

### Over-represented seed descriptors

```{r keyness-seed}
pass1 <- corpus_meta |>
  filter(pmid %in% pmids) |>
  mesh_keyness(min_count = 3L)

pass1 |>
  filter(direction == "over") |>
  arrange(desc(z)) |>
  select(DescriptorName, corpus_count, corpus_prop, baseline_prop, log_odds, z) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  head(15) |>
  DT::datatable(rownames = FALSE, options = list(scrollX = TRUE))
```

Recent papers may not be MeSH-indexed yet and the baseline is a snapshot; keyness describes the corpus, it does not judge relevance.

## Choosing a direction

`direction` matches the search intent: `"cited"` finds shared foundational references, `"citing"` finds later work building on the corpus, and `"both"` casts the widest net.

```{r direction}
backward <- citation_snowball(seed_icites, direction = "cited",  min_links = 2)
forward  <- citation_snowball(seed_icites, direction = "citing", min_links = 2)

data.frame(
  direction    = c("cited (foundational)", "citing (downstream)"),
  n_candidates = c(sum(!backward$seed), sum(!forward$seed))
)
```

## Iterating

Re-seed by feeding the expanded PMIDs back through the iCite endpoint and snowballing again; each hop keeps the same audit columns.

```{r reseed, eval=FALSE}
hop2 <- snowball$pmid |>
  get_records(endpoint = "icites", cores = 1L, sleep = 0.25) |>
  citation_snowball(direction = "both", min_links = 3, max_nodes = 500)
```

## Summary

`citation_snowball()` turns an iCite response into a ranked, auditable candidate set: it finds citation-adjacent papers a keyword query misses, the audit columns document why each was admitted, and MeSH keyness against `data_mesh_frequencies` characterizes the expansion space. It complements keyword search and manual screening rather than replacing them.
