---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
```

Two functions handle most `puremoe` workflows: `search_pubmed()` fetches PMIDs from PubMed based on a query string; `get_records()` retrieves data from a named endpoint for those PMIDs. This vignette walks through all six endpoints on a corpus of articles about political ideology.

Load packages.

```{r libs}
library(puremoe)
library(data.table)
library(DT)
```

## Step 1 -- search PubMed

`search_pubmed()` accepts standard PubMed query syntax -- field tags (`[TiAb]`, `[MeSH Terms]`, `[DP]`), Boolean operators, wildcards -- and returns a character vector of PMIDs.

```{r search}
pmids <- puremoe::search_pubmed('("political ideology"[TiAb])')
length(pmids)
```

Work with a subset for this vignette.

```{r subset}
pmids_sub <- head(pmids, 60L)
```

## Step 2 -- abstracts

`endpoint = "pubmed_abstracts"` returns one row per article: PMID, title, abstract, journal, year, authors, and MeSH terms assigned by indexers.

```{r abstracts}
abstracts <- puremoe::get_records(
  pmids_sub,
  endpoint = "pubmed_abstracts",
  cores    = 1L,
  sleep    = 0.5
)

abstracts[, pmid := as.character(pmid)]
```

```{r abstracts-table}
abstracts[1:10, .(pmid, year, journal)] |>
  DT::datatable(rownames = FALSE)
```

## Step 3 -- bibliometrics

`endpoint = "icites"` pulls citation metrics from NIH's [iCite](https://icite.od.nih.gov/): raw citation count, relative citation ratio (RCR), NIH percentile, and field-normalized citation rate.

```{r icites}
icites <- puremoe::get_records(
  pmids_sub,
  endpoint = "icites",
  cores    = 1L,
  sleep    = 0.25
)

icites[, pmid := as.character(pmid)]
```

Merge with abstracts and rank by citations.

```{r merge}
dt <- merge(
  abstracts[, .(pmid, year, journal, articletitle)],
  icites[, .(pmid, citation_count, relative_citation_ratio, nih_percentile)],
  by  = "pmid",
  all = TRUE
)

setorder(dt, -citation_count)

dt[1:10, .(pmid, year, citation_count, relative_citation_ratio)] |>
  DT::datatable(rownames = FALSE)
```

## Step 4 -- affiliations

`endpoint = "pubmed_affiliations"` returns one row per author–affiliation pair -- useful for mapping institutional and geographic reach.

```{r affiliations}
affiliations <- puremoe::get_records(
  head(pmids_sub, 30L),
  endpoint = "pubmed_affiliations",
  cores    = 1L,
  sleep    = 0.5
)
```

```{r affiliations-table}
affiliations[1:10, .(pmid, Author, Affiliation)] |>
  DT::datatable(rownames = FALSE)
```

A quick look at country representation across affiliations.

```{r affiliation-country}
affiliations[, aff_lower := tolower(Affiliation)]

data.table(
  country = c("USA", "UK", "Canada", "Germany", "Netherlands", "Australia"),
  n = c(
    affiliations[grepl("usa|united states|u\\.s\\.", aff_lower), .N],
    affiliations[grepl("united kingdom|\\buk\\b|england", aff_lower), .N],
    affiliations[grepl("canada", aff_lower), .N],
    affiliations[grepl("germany|deutschland", aff_lower), .N],
    affiliations[grepl("netherlands", aff_lower), .N],
    affiliations[grepl("australia", aff_lower), .N]
  )
)[n > 0][order(-n)]
```

## Step 5 -- named entities

`endpoint = "pubtations"` queries [PubTator3](https://www.ncbi.nlm.nih.gov/research/pubtator3/) and returns annotated entities -- genes, diseases, chemicals, species, mutations, cell lines -- identified in titles and abstracts.

```{r pubtations}
pubtations <- puremoe::get_records(
  head(pmids_sub, 40L),
  endpoint = "pubtations",
  cores    = 1L
)
```

Entity-type counts and top mentions.

```{r pubtations-types}
pubtations[!is.na(type), .N, by = type][order(-N)]
```

```{r pubtations-top}
pubtations[!is.na(type), .N, by = .(type, text)][order(-N)] |>
  head(15) |>
  DT::datatable(rownames = FALSE)
```

## Step 6 -- references

`endpoint = "pubmed_references"` returns cited references per article -- one row per reference -- useful for citation network construction or identifying foundational literature.

```{r references}
refs <- puremoe::get_records(
  head(pmids_sub, 20L),
  endpoint = "pubmed_references",
  cores    = 1L,
  sleep    = 0.5
)

refs[1:10, .(pmid, cited_pmid, cited_doi, citation)] |>
  DT::datatable(rownames = FALSE)
```

## Endpoint schemas

`endpoint_info()` lists all available endpoints; pass an endpoint name for column definitions, rate limits, and notes.

```{r endpoint-info}
puremoe::endpoint_info()
```

```{r endpoint-detail}
puremoe::endpoint_info("pubmed_abstracts")
```

## Summary

`search_pubmed()` → `get_records()` is the standard retrieval path. All six endpoints return plain `data.table` objects and compose naturally with `|>`. Use `cores > 1` and an NCBI API key for faster retrieval on large PMID sets.
