---
title: "Getting Started with puremoe"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with puremoe}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

The `puremoe` package provides a unified interface for retrieving data from
multiple PubMed/NLM databases. Two functions handle most workflows:
`search_pubmed()` fetches PMIDs matching a query; `get_records()` retrieves
data from a named endpoint for those PMIDs.

```{r library}
library(puremoe)
library(data.table)
```

---

## Search PubMed

`search_pubmed()` accepts standard PubMed query syntax, including field tags
such as `[TiAb]` (title/abstract), `[MeSH Terms]`, and `[DP]` (date
published). It returns a character vector of PMIDs.

```{r search}
pmids <- puremoe::search_pubmed('("political ideology"[TiAb])')
length(pmids)
head(pmids)
```

For this vignette we work with a manageable subset.

```{r subset}
pmids_sub <- head(pmids, 60L)
```

---

## Abstracts and metadata

`endpoint = "pubmed_abstracts"` returns one row per article with bibliographic
fields: PMID, title, abstract text, journal, year, authors, and MeSH terms.

```{r abstracts}
abstracts <- puremoe::get_records(
  pmids_sub,
  endpoint = "pubmed_abstracts",
  cores    = 1L,
  sleep    = 0.5
)

abstracts[1:3, .(pmid, year, journal, articletitle)]
```

```{r abstracts-cols}
names(abstracts)
```

---

## Bibliometric data via iCite

`endpoint = "icites"` pulls citation metrics from NIH's
[iCite](https://icite.od.nih.gov/) database: raw citation counts, relative
citation ratio (RCR), field-citation rate, and open-access status.

```{r icites}
icites <- puremoe::get_records(
  pmids_sub,
  endpoint = "icites",
  cores    = 1L,
  sleep    = 0.25
)

icites[1:3, .(pmid, citation_count, relative_citation_ratio, nih_percentile, is_clinical)]
```

Merge with abstracts for a single working table:

```{r merge}
dt <- merge(
  abstracts[, .(pmid, year, journal, articletitle)],
  icites[, .(pmid, citation_count, relative_citation_ratio, nih_percentile)],
  by  = "pmid",
  all = TRUE
)

setorder(dt, -citation_count)
head(dt[, .(pmid, year, citation_count, relative_citation_ratio, nih_percentile)])
```

---

## Author affiliations

`endpoint = "pubmed_affiliations"` returns one row per author–affiliation
pair, useful for identifying institutional and geographic patterns.

```{r affiliations}
affiliations <- puremoe::get_records(
  head(pmids_sub, 30L),
  endpoint = "pubmed_affiliations",
  cores    = 1L,
  sleep    = 0.5
)

affiliations[1:5, .(pmid, Author, Affiliation)]
```

A simple look at country representation:

```{r affiliation-country}
affiliations[, aff_lower := tolower(Affiliation)]

country_counts <- data.table(
  country = c("USA", "UK", "Canada", "Germany", "Netherlands", "Australia"),
  n = c(
    affiliations[grepl("usa|united states|u\\.s\\.", aff_lower), .N],
    affiliations[grepl("united kingdom|\\buk\\b|england", aff_lower), .N],
    affiliations[grepl("canada", aff_lower), .N],
    affiliations[grepl("germany|deutschland", aff_lower), .N],
    affiliations[grepl("netherlands", aff_lower), .N],
    affiliations[grepl("australia", aff_lower), .N]
  )
)

country_counts[n > 0][order(-n)]
```

---

## Named-entity annotations via PubTator3

`endpoint = "pubtations"` queries
[PubTator3](https://www.ncbi.nlm.nih.gov/research/pubtator3/) and returns
annotated entities (genes, diseases, chemicals, species, mutations, cell lines)
found in titles and abstracts.

```{r pubtations}
pubtations <- puremoe::get_records(
  head(pmids_sub, 40L),
  endpoint = "pubtations",
  cores    = 1L
)

pubtations[1:8, .(pmid, type, text, identifier)]
```

Entity-type breakdown:

```{r pubtations-types}
pubtations[!is.na(type), .N, by = type][order(-N)]
```

Most frequently mentioned entities:

```{r pubtations-top}
pubtations[!is.na(type), .N, by = .(type, text)][order(-N)] |> head(12)
```

---

## References

`endpoint = "pubmed_references"` returns cited references for each article —
useful for citation network construction or identifying foundational literature.

```{r references}
refs <- puremoe::get_records(
  head(pmids_sub, 20L),
  endpoint = "pubmed_references",
  cores    = 1L,
  sleep    = 0.5
)

refs[1:6, .(pmid, ref_pmid, ref_citation)]
```

---

## Endpoint metadata

`endpoint_info()` lists available endpoints and their schemas — handy when
building applications or checking column definitions.

```{r endpoint-info}
puremoe::endpoint_info()
```

```{r endpoint-detail}
puremoe::endpoint_info("pubmed_abstracts")
```

---

## Session info

```{r session}
sessionInfo()
```
