---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
```

`puremoe` provides a unified interface to PubMed and the wider NIH/NLM data stack, organized as two layers: **retrieve** records from public services with `search_pubmed()` and `get_records()`, then **analyze** the returned tables locally with no further API calls. PMIDs are the common key throughout. This vignette covers the retrieval layer end-to-end; the analysis layer is covered in the topic vignettes linked at the close.

```{r libs}
library(puremoe)
library(dplyr)
library(DT)
```

## Search

`search_pubmed()` accepts standard PubMed query syntax and returns a vector of PMIDs.

```{r search}
pmids <- puremoe::search_pubmed('("political ideology"[TiAb])')
length(pmids)
```

## Abstracts

```{r abstracts}
abstracts <- puremoe::get_records(
  pmids,
  endpoint = "pubmed_abstracts",
  cores    = 1L,
  sleep    = 0.5
)
```

### Abstract records

```{r abstracts-table}
abstracts |>
  select(pmid, year, journal, articletitle) |>
  head(10) |>
  DT::datatable(rownames = FALSE, options = list(scrollX = TRUE))
```

## Annotations

The `annotations` column is a list of per-article data frames containing MeSH terms, chemical names, and keywords.

### Annotation terms

```{r annotations}
bind_rows(abstracts$annotations) |>
  filter(!is.na(DescriptorName), nzchar(DescriptorName)) |>
  head(20) |>
  DT::datatable(rownames = FALSE, options = list(scrollX = TRUE))
```

## Affiliations

```{r affiliations}
affiliations <- puremoe::get_records(
  pmids,
  endpoint = "pubmed_affiliations",
  cores    = 1L,
  sleep    = 0.5
)
```

### Author affiliations

```{r affiliations-display}
affiliations |>
  head(15) |>
  DT::datatable(rownames = FALSE, options = list(scrollX = TRUE))
```

## iCite metrics

iCite lags PubMed indexing; the most recent PMIDs in a search may not yet have citation metrics.

```{r icites}
icites <- puremoe::get_records(
  pmids,
  endpoint = "icites",
  cores    = 1L,
  sleep    = 0.25
)
```

### Indexed records

```{r icites-display}
icites |>
  filter(!is.na(citation_count)) |>
  mutate(field_citation_rate = round(field_citation_rate, 3)) |>
  select(-citation_net, -cited_by_clin) |>
  head(10) |>
  DT::datatable(rownames = FALSE, options = list(scrollX = TRUE))
```

## PubTator annotations

```{r pubtator}
pubtator <- puremoe::get_records(
  pmids,
  endpoint = "pubtator",
  cores    = 1L
)
```

### Entity mentions

```{r pubtator-display}
pubtator$entities |>
  select(-passage_text, -passage_offset) |>
  head(15) |>
  DT::datatable(rownames = FALSE, options = list(scrollX = TRUE))
```

## Full text

Full-text retrieval requires open-access PMC articles. `pmid_to_ftp()` resolves
PMIDs to XML URLs via the PMC Cloud Service on AWS S3, filtering to only those
with open-access full text available. In August 2026, NCBI will complete its
migration from the legacy PMC FTP Service to the Cloud Service; `puremoe` already
uses the new service.

```{r ftp}
ftp <- puremoe::pmid_to_ftp(pmids = head(pmids, 25L))
```

### Open-access URLs

```{r ftp-display}
ftp |>
  head(5L) |>
  DT::datatable(rownames = FALSE, options = list(scrollX = TRUE))
```

### Section-level text

```{r fulltext}
fulltext <- puremoe::get_records(
  head(ftp$url, 5L),
  endpoint = "pmc_fulltext",
  cores    = 1L
)

fulltext |>
  mutate(text = sapply(strsplit(text, "\\s+"), function(w) paste0(paste(head(w, 15), collapse = " "), "..."))) |>
  DT::datatable(rownames = FALSE, options = list(scrollX = TRUE))
```

## Endpoint schemas

`endpoint_info()` returns column definitions, rate limits, and notes for any endpoint.

```{r endpoint-info}
puremoe::endpoint_info()
```

```{r endpoint-detail}
puremoe::endpoint_info("icites")
```

## Next steps

The tables retrieved here feed `puremoe`'s local analysis layer -- transforms over
the same data frames, with no further API calls:

- **MeSH tables** -- look up descriptors, navigate the hierarchy, and use the
  PubMed-wide frequency baseline.
- **Citation snowballing** -- expand a seed corpus along `icites` citation links
  and characterize the result with MeSH keyness.
- **PubTator context and relation networks** -- anchor entity mentions to
  sentences, count co-occurrence, and build relation networks with evidence.
