---
title: "MeSH-Guided Search"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MeSH-Guided Search}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
```

PubMed indexes articles with Medical Subject Headings (MeSH) -- a controlled vocabulary maintained by NLM. A `[MeSH Terms]` query retrieves articles explicitly tagged with a concept regardless of how the authors phrased it, and triggers *explosion*: all narrower terms in the hierarchy are included automatically. `puremoe` ships two MeSH reference datasets, the thesaurus and the tree, that make it straightforward to identify the right headings before running a search.

Load packages.

```{r libs}
library(puremoe)
library(data.table)
library(DT)
```

## MeSH thesaurus

`data_mesh_thesaurus()` downloads and combines the MeSH Descriptor Thesaurus and Supplementary Concept Records (SCR). It returns one row per term, including all synonyms and entry terms for each descriptor.

```{r thesaurus}
thesaurus <- puremoe::data_mesh_thesaurus()
dim(thesaurus)
names(thesaurus)
```

```{r thesaurus-head}
head(thesaurus, 6) |> DT::datatable(rownames = FALSE)
```

### Finding terms related to a topic

Search the thesaurus for headings related to *political ideology*.

```{r thesaurus-search}
hits <- thesaurus[grepl("politi", TermName, ignore.case = TRUE)]
hits[, .(DescriptorName, TermName)] |>
  unique() |>
  DT::datatable(rownames = FALSE)
```

Pull out the canonical descriptor names -- one row per MeSH heading.

```{r thesaurus-descriptors}
descriptors <- hits[, .(DescriptorName)] |> unique()
descriptors
```

## MeSH trees

`data_mesh_trees()` provides the hierarchical classification structure. Each descriptor can appear in multiple branches; tree numbers encode the full path (e.g., `I01.880` = Social Sciences > Political Science).

```{r trees}
trees <- puremoe::data_mesh_trees()
dim(trees)
names(trees)
```

### Locating descriptors in the hierarchy

Find where our candidate descriptors sit.

```{r trees-politics}
political_trees <- trees[DescriptorName %in% descriptors$DescriptorName]
political_trees[, .(DescriptorName, TreeNumber)] |>
  unique() |>
  DT::datatable(rownames = FALSE)
```

### Exploring a branch

Retrieve everything in the `I01.880` subtree (Political Science and related concepts).

```{r trees-branch}
branch <- trees[grepl("^I01\\.880", TreeNumber)]
branch[, .(DescriptorName, TreeNumber)][order(TreeNumber)] |>
  unique() |>
  head(20) |>
  DT::datatable(rownames = FALSE)
```

### Siblings and parent context

Look one level up to see what else sits under `I01` (Social Sciences).

```{r trees-siblings}
siblings <- trees[grepl("^I01\\.[0-9]+$", TreeNumber)]
siblings[, .(DescriptorName, TreeNumber)][order(TreeNumber)] |>
  unique() |>
  DT::datatable(rownames = FALSE)
```

## Building a MeSH query

Construct a controlled-vocabulary PubMed query from the descriptors identified above. `[MeSH Terms]` triggers explosion; use `[MeSH:noexp]` to suppress it.

```{r query-build}
mesh_terms  <- c("Politics", "Political Systems")
mesh_query  <- paste(sprintf('"%s"[MeSH Terms]', mesh_terms), collapse = " OR ")
cat(mesh_query)
```

Combine with a free-text component to capture articles not yet indexed or tagged under a different heading.

```{r query-combined}
combined_query <- paste0(
  '(', mesh_query, ')',
  ' OR ',
  '("political ideology"[TiAb])'
)
cat(combined_query)
```

## Comparing queries

Run both queries and compare recall.

```{r search-compare}
pmids_mesh     <- puremoe::search_pubmed(mesh_query)
pmids_combined <- puremoe::search_pubmed(combined_query)

data.table(
  query         = c("MeSH only", "Combined (MeSH + free-text)"),
  n_pmids       = c(length(pmids_mesh), length(pmids_combined))
)
```

```{r search-overlap}
cat("Unique to free-text:  ", length(setdiff(pmids_combined, pmids_mesh)), "\n")
cat("Unique to MeSH:       ", length(setdiff(pmids_mesh, pmids_combined)), "\n")
cat("In both:              ", length(intersect(pmids_mesh, pmids_combined)), "\n")
```

## Retrieving records

Fetch abstracts for a subset of the combined result set and inspect which MeSH headings PubMed has assigned.

```{r abstracts}
pmids_sub <- head(pmids_combined, 40L)

abstracts <- puremoe::get_records(
  pmids_sub,
  endpoint = "pubmed_abstracts",
  cores    = 1L,
  sleep    = 0.5
)
```

```{r mesh-assigned}
assigned_mesh <- abstracts[
  !is.na(mesh_terms),
  .(mesh = unlist(mesh_terms)),
  by = pmid
]

assigned_mesh[, .N, by = mesh][order(-N)] |>
  head(20) |>
  DT::datatable(rownames = FALSE)
```

## Persistent storage

Both datasets are ~10 MB. Download once and cache locally to avoid re-fetching each session.

```{r persistent, eval=FALSE}
thesaurus <- puremoe::data_mesh_thesaurus(use_persistent_storage = TRUE)
trees     <- puremoe::data_mesh_trees(use_persistent_storage = TRUE)
```

## Summary

`data_mesh_thesaurus()` → find descriptors → `data_mesh_trees()` → explore hierarchy → build a `[MeSH Terms]` query → `search_pubmed()` → `get_records()`. MeSH explosion means a single well-chosen heading often retrieves more relevant records than a list of free-text synonyms.
