---
title: "MeSH-Guided Search with puremoe"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MeSH-Guided Search with puremoe}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

PubMed indexes articles with **Medical Subject Headings (MeSH)** — a controlled
vocabulary maintained by NLM. MeSH terms are more precise than free-text
searches: a `[MeSH Terms]` query retrieves articles that have been
explicitly tagged with a concept, regardless of how the authors phrased it.

`puremoe` bundles two MeSH reference datasets: the **thesaurus** (descriptors
+ supplementary concepts + synonyms) and the **tree** (the hierarchical
classification structure). This vignette shows how to use both to build
targeted PubMed queries and understand the shape of a literature.

```{r library}
library(puremoe)
library(data.table)
```

---

## MeSH thesaurus

`data_mesh_thesaurus()` downloads and combines the MeSH Descriptor Thesaurus
and the Supplementary Concept Record (SCR) thesaurus. It returns a `data.table`
with one row per term (including synonyms/entry terms for each descriptor).

```{r thesaurus}
thesaurus <- puremoe::data_mesh_thesaurus()
dim(thesaurus)
names(thesaurus)
```

```{r thesaurus-head}
head(thesaurus, 4)
```

### Finding terms related to a topic

Search the thesaurus for concepts related to *political* or *ideology*:

```{r thesaurus-search}
hits <- thesaurus[grepl("politi", TermName, ignore.case = TRUE)]
hits[, .(DescriptorName, TermName)] |> unique() |> head(20)
```

Narrow to the canonical descriptor names only (one row per MeSH heading):

```{r thesaurus-descriptors}
descriptors <- hits[, .(DescriptorName)] |> unique()
descriptors
```

---

## MeSH trees

`data_mesh_trees()` provides the hierarchical tree structure. Each MeSH
descriptor can appear in multiple branches; tree numbers encode the path
(e.g., `I01.880` = Social Sciences > Political Science).

```{r trees}
trees <- puremoe::data_mesh_trees()
dim(trees)
names(trees)
```

```{r trees-head}
head(trees, 4)
```

### Locating descriptors in the hierarchy

Find tree positions for our candidate descriptors:

```{r trees-politics}
political_trees <- trees[DescriptorName %in% descriptors$DescriptorName]
political_trees[, .(DescriptorName, TreeNumber)] |> unique()
```

### Exploring a branch

Pick a tree prefix and retrieve everything in that subtree. Here we look at
the **I01.880** branch (Political Science and related):

```{r trees-branch}
branch <- trees[grepl("^I01\\.880", TreeNumber)]
branch[, .(DescriptorName, TreeNumber)][order(TreeNumber)] |> unique() |> head(20)
```

### Finding parent and sibling descriptors

Look one level up from `I01.880` to see what else sits in `I01` (Social
Sciences):

```{r trees-siblings}
siblings <- trees[grepl("^I01\\.[0-9]+$", TreeNumber)]
siblings[, .(DescriptorName, TreeNumber)][order(TreeNumber)] |> unique()
```

---

## Building a MeSH-based PubMed query

With the descriptors in hand we can construct a controlled-vocabulary query.
`[MeSH Terms]` triggers MeSH **explosion** (includes all narrower terms in the
hierarchy); `[MeSH:noexp]` disables this if you want exact-match only.

```{r query-build}
mesh_terms <- c("Politics", "Political Systems")

# Collapse to a single OR-joined PubMed query
mesh_query <- paste(
  sprintf('"%s"[MeSH Terms]', mesh_terms),
  collapse = " OR "
)
cat(mesh_query)
```

Combine with a free-text component to capture articles not yet indexed or
indexed under a different heading:

```{r query-combined}
combined_query <- paste0(
  '(', mesh_query, ')',
  ' OR ',
  '("political ideology"[TiAb])'
)
cat(combined_query)
```

---

## Searching and comparing

Run both the pure MeSH and combined queries to see the difference in recall:

```{r search-mesh}
pmids_mesh     <- puremoe::search_pubmed(mesh_query)
pmids_combined <- puremoe::search_pubmed(combined_query)

cat("MeSH-only PMIDs:    ", length(pmids_mesh), "\n")
cat("Combined PMIDs:     ", length(pmids_combined), "\n")
cat("Free-text adds:     ", length(setdiff(pmids_combined, pmids_mesh)), "\n")
cat("MeSH adds:          ", length(setdiff(pmids_mesh, pmids_combined)), "\n")
```

---

## Retrieving and inspecting records

Fetch abstracts for a subset of the combined result set and check which MeSH
terms PubMed has actually assigned:

```{r abstracts}
pmids_sub <- head(pmids_combined, 40L)

abstracts <- puremoe::get_records(
  pmids_sub,
  endpoint = "pubmed_abstracts",
  cores    = 1L,
  sleep    = 0.5
)
```

```{r mesh-assigned}
# MeSH terms are stored as a list-column; unnest and count
assigned_mesh <- abstracts[
  !is.na(mesh_terms),
  .(mesh = unlist(mesh_terms)),
  by = pmid
]

assigned_mesh[, .N, by = mesh][order(-N)] |> head(15)
```

---

## Persistence across sessions

Both datasets are ~10 MB. If you work with them regularly, store them once
and reload on future sessions rather than re-downloading each time:

```{r persistent, eval = FALSE}
# Download once to a persistent location
thesaurus <- puremoe::data_mesh_thesaurus(use_persistent_storage = TRUE)
trees     <- puremoe::data_mesh_trees(use_persistent_storage = TRUE)

# Subsequent calls find the cached files immediately
thesaurus <- puremoe::data_mesh_thesaurus(use_persistent_storage = TRUE)
```

---

## Session info

```{r session}
sessionInfo()
```
