---
title: "MeSH Tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MeSH Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
```

`puremoe` ships two MeSH reference tables: a thesaurus of descriptors and entry terms, and a tree of hierarchical classifications.

```{r libs}
library(puremoe)
library(dplyr)
library(DT)
```

## MeSH thesaurus

`data_mesh_thesaurus()` downloads and combines the MeSH Descriptor Thesaurus and Supplementary Concept Records (SCR). One row per term, including synonyms and entry terms for each descriptor.

```{r thesaurus}
thesaurus <- puremoe::data_mesh_thesaurus()
```

```{r thesaurus-table}
thesaurus |>
  head(20) |>
  DT::datatable(rownames = FALSE, options = list(scrollX = TRUE))
```

## MeSH trees

`data_mesh_trees()` provides the hierarchical classification structure. Each descriptor can appear in multiple branches; `tree_location` encodes the full path (e.g., `I01.880.604` = Social Sciences > Political Science > Political Systems).

```{r trees}
trees <- puremoe::data_mesh_trees()
```

```{r trees-table}
trees |>
  head(20) |>
  DT::datatable(rownames = FALSE, options = list(scrollX = TRUE))
```
