---
title: "PubTator Context, Co-occurrence, and Networks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PubTator Context, Co-occurrence, and Networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`puremoe` retrieves PubTator3 entity mentions and relation pairs with one call, then reshapes them locally into sentence context, co-occurrence counts, entity networks, and the sentences behind extracted relations. Everything after the initial fetch is a local transform on the retrieved tables and makes no further API calls.

We use a biomedical corpus with rich entity and relation annotation throughout.

```{r pubtator-fetch, eval=FALSE}
library(puremoe)
library(data.table)

pmids <- search_pubmed('"doxorubicin"[TiAb] AND "cardiotoxicity"[TiAb]')

pt <- get_records(head(pmids, 25L), endpoint = "pubtator", cores = 1)

names(pt)
pt$entities
pt$relations
```



## Sentence context

`pubtator_context()` preserves PubTator entity spans, adds sentence IDs and sentence-relative spans to entities, adds readable entity labels and sentence anchors to relations, and returns a sentence lookup table.

```{r pubtator-context, eval=FALSE}
ctx <- pubtator_context(pt)

ctx$entities
ctx$relations
ctx$sentences
```

## Co-occurrence counts

Use `pubtator_cooccurrence()` for count-only summaries over sentence windows. The function accepts either the full context object or `ctx$entities`. `window = 0` counts entities in the same sentence; larger windows reach across adjacent sentences within the same passage.

```{r cooccurrence, eval=FALSE}
# entity-type pairs in the same sentence
pubtator_cooccurrence(ctx, window = 0, by = "type")

# specific entity pairs, within one sentence of each other
pubtator_cooccurrence(ctx, window = 1, by = "entity")
```

## Relation networks

`pubtator_network()` turns contextualized PubTator relations into a directed entity network plus a lean evidence table. Nodes are specific entities keyed on their PubTator `identifier` when present, falling back to `type:text`; edges are directed PubTator relation assertions; evidence rows map each edge back to `pmid`, `relation_id`, and the shared sentence when the endpoint mentions occur in the same sentence.

```{r network-relations, eval=FALSE}
rel_net <- pubtator_network(ctx)

rel_net
```

`puremoe` takes on no graphing dependency itself, but the graph tables pass straight to a graph package such as igraph:

```{r network-igraph, eval=FALSE}
igraph::graph_from_data_frame(rel_net, vertices = rel_net)
```

## Relation evidence

`relation_evidence()` aggregates PubTator3 relations into a per-relation summary: one row per unique relationship, reporting how many documents (`n_pmids`) and sentences (`n_sentences`) assert it, with the supporting sentences kept in a `sentences` list-column for reading. The available relation types depend on the corpus, so inspect them first.

```{r relation-types, eval=FALSE}
sort(table(ctx$relations$relation_type), decreasing = TRUE)
```

Supplying an `icites` table adds a `citation_support` column (citation counts summed over the distinct supporting papers) and ranks the relations by it, so the best-attested assertions surface first.

```{r relation-evidence, eval=FALSE}
icites <- get_records(head(pmids, 25L), endpoint = "icites", cores = 1)

ev <- relation_evidence(ctx, 
                        relation_type = "positive_correlation", 
                        icites = icites)
ev

# read the sentences behind the top-ranked relation
ev$sentences[[1]]
```

This turns a flat relation table into a ranked picture of which relationships the corpus actually supports, each row traceable through `sentences` and `pmids` back to its source. Pass `relation_type = NULL` (the default) to summarize every type.
