---
title: "puremoe data sets"
---

`puremoe` includes several MeSH-/PMC-based tables. These resources are too large for a CRAN package.  Instead, the functions below download data from a simple GitHub library at [mesh-resources](https://github.com/jaytimm/mesh-resources).  Data sets can be cached in the package data folder by setting the `use_persistent_storage` to TRUE.  Otherwise you can just re-download to a temp directory each time functions are called.


https://r-pkgs.org/vignettes.html

## MeSH Resources


### MeSH thesaurus

```{r message=FALSE, warning=FALSE}
thesaurus <- puremoe::data_mesh_thesaurus(use_persistent_storage = T)
```


```{r}
thesaurus |> head() |> 
  DT::datatable(rownames = F, options = list(scrollX = TRUE))
```


### MeSH Trees

```{r message=FALSE, warning=FALSE}
trees <- puremoe::data_mesh_trees(use_persistent_storage = T)
```


```{r}
trees |> head() |> DT::datatable(rownames = F, options = list(scrollX = TRUE))
```


### MeSH Pharmacological Actions

```{r message=FALSE, warning=FALSE}
pharm_actions <- puremoe::data_pharm_action(use_persistent_storage = T)
```



```{r}
pharm_actions |> head() |> DT::datatable(rownames = F, options = list(scrollX = TRUE))
```


### MeSH Term Embeddings

```{r message=FALSE, warning=FALSE}
embeddings <- puremoe::data_mesh_embeddings(use_persistent_storage = T)
```


```{r}
textpress::sem_nearest_neighbors(x = 'Aripiprazole', matrix = embeddings) |>
  DT::datatable(rownames = F, options = list(scrollX = TRUE))
```




## Pubmed Central

```{r message=FALSE, warning=FALSE}
pmc_list <- puremoe::data_pmc_list(use_persistent_storage = T)
```


```{r}
pmc_list |> head() |> DT::datatable(rownames = F, options = list(scrollX = TRUE))
```







